---
title: "NCI Biomarker Data Explorer"
author: "Fred Hutch Data Science Lab"
format: dashboard
server: shiny
---

```{r}
#| context: setup
library(tidyverse)
library(purrr)
library(glue)
library(here)
library(ggplot2)

biomarker_data <- read_csv(
  here("data", "fhcc_trial_biomarkers_tidy.csv")
) |>
  rename(biomarker_name = name)

disease_type <- read_rds(
  here("data", "fhcc_trial_disease_top.rds")
) |>
  rename(disease = name)

fhcc_trials_nci <- fhcc_trials |>
  filter(!is.na(nci_trial_id))

disease_main <- disease_type |>
  filter(type == "maintype") |>
  filter(!str_detect(disease, "Other"),
         disease != "Solid Tumor")

biomarker_disease <- biomarker_data |>
  left_join(disease_main, by = c("trial" = "nci_id")) 
```

# {.sidebar}

```{r}
selectInput(
  'disease', 'Disease Type', 
  sort(unique(biomarker_disease$disease))
  )
```

# Plot

```{r}
plotOutput('barPlot')
```

# Data

```{r}
DT::DTOutput('trialTable')
```


```{r}
#| context: server

dataset <- reactive({
  biomarker_disease |> 
    filter(disease == input$disease)
  
})

output$barPlot <- renderPlot({

  plot_df <- dataset() |>
    filter(eligibility_criterion == "inclusion") |>
    distinct(trial, biomarker_name) |>
    group_by(biomarker_name) |>
    summarise(biomarker_count = n()) |>
    ungroup()

  p <- ggplot(
    plot_df,
    aes(x = reorder(biomarker_name, biomarker_count), y = biomarker_count)
  ) +
    geom_col(fill = "#0A799A") +
    scale_y_continuous(breaks = scales::breaks_pretty()) +
    coord_flip() +
    theme_bw() +
    theme(legend.position = "none",
          plot.title.position = "plot",
          plot.title = ggtext::element_textbox_simple()) +
    labs(
      x = "Biomarker name",
      y = "Count of NCI-sponsored trials",
      title = glue("Count of Fred Hutch NCI-Sponsored Trials \n for {input$disease} with a Molecular Biomarker Used in Inclusion Criteria, by Biomarker Name")
    )
  p
})

output$trialTable <- DT::renderDT(
  
  dataset() |>
    filter(eligibility_criterion == "inclusion") |>
    distinct(trial, brief_title, biomarker_name, code) |>
    select(
      `NCI ID` = trial,
      Title = brief_title,
      Biomarker = biomarker_name,
      `Concept ID` = code
    ),
  filter = "top"
  
)

```


# Background

The National Cancer Institute (NCI) maintains a database of [NCI-sponsored clinical trials](https://www.cancer.gov/research/participate/clinical-trials-search) and their associated metadata. We used the [NCI API](https://clinicaltrialsapi.cancer.gov/doc) to pull biomarker and disease concepts for eachclinical trial. Each biomarker and disease concept has a concept ID and is part of a standardized terminology in the NCI thesaurus ([CTRP Molecular Genetic Biomarker Terminology](https://nciterms.nci.nih.gov/ncitbrowser/ConceptReport.jsp?dictionary=NCI_Thesaurus&version=24.03d&code=C142800&ns=ncit)). The [clinicaltrials.gov site also has an API](https://clinicaltrials.gov/data-api/about-api), but it does not include biomarker concepts.

## Source Data

Data were pulled from the NCI API on December 12, 2024. There were 1876 trials in the system that listed Fred Hutch as a trial site. Of these, `r length(unique(biomarker_data$trial))` had a biomarker associated with the trial inclusion/exclusion criteria. There are `r length(unique(biomarker_data$biomarker_name))` distinct biomarkers associated with these trials.

